---
title: "Monthly albedo comparison of CMIP and ERA datasets"
author: "Ben Müller"
date: "8. Oktober 2015"
output:
  html_document:
    keep_md: yes
---

## Analysis

### Basic calculations for CMIP and ERA-Interim data

Setting the operating directories for CMIP and ERA-interim data ...

```{r,echo=T,message=F,warning=F,tidy=TRUE}
##### Operating directories #####

WD.CMIP="MPI-M.CMIP5.AMIP\\Amon\\r3i1p1"

WD.ERA="ERA-Interim"
```

and run *source('Main.R')* will start the preprocessing, including:

* loading monthly CMIP data for upwelling and downwelling short wave data,
* calculating albedo from the ratio of shortwave data,
* if necessary, aggregating daily ERA-Interim albedo data at 12:00 am to monthly data,
* loading monthly ERA-Interim albedo data,
* adjusting the resolution of the finer ERA-Interim data to the coarser CMIP data,
* masking data that is NA in one of the datasets, and
* wraping data to center the maps on the Atlantic ocean.

```{r,echo=F,message=F,warning=F,tidy=T,comment=""}
# source the preprocessing
source('Main.R')
```
  
***
  
### Overview on the datasets and their differences

```{r,echo=F,message=F,warning=F,fig.width=9.2/2,fig.show='hold'}
# plot overview
spplot(mean(Sets$CMIP),at=seq(0,1,l=21),colorkey=list(space="bottom"),scales = list(tick.number=5,draw = TRUE),col.regions=colorRampPalette(c("gray11","lightgoldenrod3","khaki1"))(21),main="CMIP mean albedo")

spplot(mean(Sets$ERA),at=seq(0,1,l=21),colorkey=list(space="bottom"),scales = list(tick.number=5,draw = TRUE),col.regions=colorRampPalette(c("gray11","lightgoldenrod3","khaki1"))(21),main="ERA mean albedo")
```

```{r,echo=F,message=F,warning=F,fig.width=9.2,fig.show='hold',fig.align='center'}
spplot(mean(Sets$CMIP)-mean(Sets$ERA),at=seq(-1,1,l=40),colorkey=list(space="bottom"),scales = list(tick.number=5,draw = TRUE),col.regions=colorRampPalette(c("darkblue","blue","grey90","red","darkred"))(40),main="difference of mean albedo (CMIP-ERA)")
```
  
***

### Spatially aggregated differences analysis in the time series of CMIP and ERA

```{r,echo=F,message=F,warning=F,fig.width=9.2,fig.show='hold',fig.align='center'}
# plot timeseries of differences

diff.df=na.omit(as.data.frame(Sets$CMIP-Sets$ERA))

DF.var=data.frame(t(data.frame(lapply(diff.df,fivenum))))
meaning=c("minimum","lower hinge","median","upper hinge","maximum")
colnames(DF.var)=meaning
DF.var=cbind(date=as.Date(paste(names(Sets[[1]]),"15",sep="."),"X%Y.%m.%d"),DF.var)

layout(matrix(
  c(
    1,1,1,1,1,1,1,1,1,
    1,1,1,1,1,1,2,2,1,
    1,1,1,1,1,1,2,2,1,
    1,1,1,1,1,1,1,1,1,
    1,1,1,1,1,1,1,1,1
  ),byrow=T,ncol=9
))

par(mar=c(5,5,3,7))
plot(DF.var$date,DF.var$median,xaxs="i",type="l",ylim=c(min(DF.var[,-1]),max(DF.var[,-1])),col="white",xlab="time [months]",ylab="d(albedo) [CMIP-ERA]",main="time series of spatially averaged monthly differences in albedo (CMIP-ERA)",cex.main=1.3,cex.lab=1.3)
polygon(c(DF.var$date,rev(DF.var$date)),c(DF.var$minimum,rev(DF.var$maximum)),col="#EE000030",border=NA)
polygon(c(DF.var$date,rev(DF.var$date)),c(DF.var$`lower hinge`,rev(DF.var$`upper hinge`)),col="#EE000070",border=NA)
lines(DF.var$date,DF.var$median,col="black")
legend("left",legend=c("median","hinges","max/min"),horiz = T,col=c("black","#EE000079","#EE000030"),lty=c(1,NA,NA),pch=c(NA,15,15),pt.cex=4,inset=0.1,cex=1.3)

DF.var.ltm=aggregate(DF.var[,-1],list(strftime(DF.var$date, "%m")),mean)
names(DF.var.ltm)[1]="date"

par(mar=c(5,4,2,1))
plot(DF.var.ltm$date,DF.var.ltm$median,xaxs="i",yaxs="i",type="l",xlim=c(1,12),ylim=0.05*c(-1,1)+c(min(DF.var.ltm[,-1]),max(DF.var.ltm[,-1])),col="white",xlab="",ylab="",main="monthly mean",cex.main=1.3,cex.lab=1.3)
polygon(c(1,12,12,1),0.05*c(-1,-1,1,1)+c(rep(min(DF.var.ltm[,-1]),2),rep(max(DF.var.ltm[,-1]),2)),col="white")
polygon(c(DF.var.ltm$date,rev(DF.var.ltm$date)),c(DF.var.ltm$minimum,rev(DF.var.ltm$maximum)),col="#EE000030",border=NA)
polygon(c(DF.var.ltm$date,rev(DF.var.ltm$date)),c(DF.var.ltm$`lower hinge`,rev(DF.var.ltm$`upper hinge`)),col="#EE000070",border=NA)
lines(DF.var.ltm$date,DF.var.ltm$median,col="black")
```
  
The time steps are aggregated in spatial means of differences and boxplot-like calculations lead to the determination of the hinges.
  
***

### Temporally differences between CMIP and ERA as maps of boxplot-like occurence

```{r,echo=F,message=F,warning=F,fig.width=9.2,fig.height=20,fig.show='hold',fig.align='center'}
# plot boxplot-like maps of time series 

fnum=calc((Sets$CMIP-Sets$ERA),fivenum)
names(fnum)=c("minimum","lower hinge","median","upper hinge","maximum")

fnum=stack(unstack(fnum)[5:1])

spplot(fnum,at=seq(-1,1,l=101),col.regions=colorRampPalette(c("darkblue","blue","grey90","red","darkred"))(100),colorkey=list(space=c("bottom")),layout = c(1, 5),scales = list(tick.number=5,draw = TRUE),main="distribution of spatial minima, maxima, boxplot-like hinge levels and median of differences between CMIP and ERA [CMIP-ERA]")
```
    
Mapped boxplot-like calculations of pixel-wise time series data of differences between CMIP and ERA.
  
***
  
### Coefficient of variation (COV) for CMIP and ERA
  
#### Time series of spatial COV

```{r,echo=F,message=F,warning=F,fig.width=9.2,fig.show='hold',fig.align='center'}
# COV for time series

sd.CMIP=unlist(lapply(lapply(unstack(Sets$CMIP),"["),sd,na.rm=T))
mean.CMIP=unlist(lapply(lapply(unstack(Sets$CMIP),"["),mean,na.rm=T))

sd.ERA=unlist(lapply(lapply(unstack(Sets$ERA),"["),sd,na.rm=T))
mean.ERA=unlist(lapply(lapply(unstack(Sets$ERA),"["),mean,na.rm=T))

DF.COV=data.frame(date=as.Date(paste(names(Sets[[1]]),"15",sep="."),"X%Y.%m.%d"),COV.CMIP=sd.CMIP/mean.CMIP,COV.ERA=sd.ERA/mean.ERA)

plot(DF.COV$date,DF.COV$COV.CMIP,xaxs="i",type="l",ylim=c(min(DF.COV[,-1]),max(DF.COV[,-1])),col="blue",lwd=2,xlab="time [months]",ylab="albedo",main="time series of spatially aggregated COV",cex.main=1.3,cex.lab=1.3)
lines(DF.var$date,DF.COV$COV.ERA,col="red",lwd=2)
legend("left",legend=c("CMIP","ERA"),horiz = T,col=c("blue","red"),lwd=c(2,2),inset=0.1,cex=1.3)
```
    
#### Maps of temporal COV

```{r,echo=F,message=F,warning=F,fig.width=9.2,fig.show='hold',fig.align='center'}
# spatial COV

sd.CMIP=calc(Sets$CMIP,sd)
mean.CMIP=calc(Sets$CMIP,mean)

sd.ERA=calc(Sets$ERA,sd)
mean.ERA=calc(Sets$ERA,mean)

spplot(stack(list(COV.CMIP=sd.CMIP/mean.CMIP,COV.ERA=sd.ERA/mean.ERA)),at=seq(0,1.5,l=101),col.regions=colorRampPalette(c("#226422","#FFFF22","#642222"))(100),colorkey=list(space=c("bottom")),scales = list(tick.number=5,draw = TRUE),main="coefficients of variations for the time series")

```
  
***
  
### Spatial aggregated temporal differences between CMIP and ERA 
  
#### By continent

```{r}
# cut differences by continents

Diff=Sets$CMIP-Sets$ERA

DF=data.frame(region=cont$REGION,)
```


#### By latitude

```{r}

```

